Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2024-05-16 manuscript.Rmd
Converted citations to csl-json, minor text updates

title: "Hawaiian reef communities are more distinct across a 90 m depth gradient than 2,500 km of shallow habitat"

@@

        - nerr

@@

  - himb: Hawai&#x02BB;i Institute of Marine Biology
  - nerr: He&#x02BB;eia National Estuarine Research Reserve

@@

  Coral reef communities are influenced by various environmental and anthropogenic factors, yet those factors have not been consistently resolved. Furthermore, due to logistical difficulties, investigations of the role of depth (particularly mesophotic coral ecosystems) in influencing coral reef community structure have been taxonomically limited. Here, we examined the effects of a suite of environmental, ecological, and anthropogenic factors on the community structure of shallow brachyuran crabs across the ~2,500 km span of the Hawaiian Archipelago. In addition, we compared shallow crab communities with those sampled on a depth transect from 12--90 m on O&#x02BB;ahu. We deployed Autonomous Reef Monitoring Structures (ARMS) on shallow reefs (12--16 m) across the latitudinal gradient of the Hawaiian Archipelago to compare with a depth gradient (12, 30, 60, and 90 m) on the south shore of O&#x02BB;ahu. We found a total of 2,284 individual crabs representing 90 morphospecies and 19 families. Shallow crab communities were significantly different between the Main and Northwest Hawaiian Islands and variation in community structure was significantly associated with six biotic, environmental, and anthropogenic factors (cumulative human impacts, chlorophyll-A, sea-surface temperature, site depth, island slope, and potential larval connectivity). In depth comparisons, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among all shallow samples across the entire archipelago. We show that 90 m of depth is a more potent determinant of brachyuran community composition than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.<!--Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems. Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to ~170 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago.-->
bibliography: references/citations.bib

@@

print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE) {
  if (!grepl("^: ",caption)) {
    caption <- paste0(": ",caption)
  }

@@

  table_str <- capture.output(pander::pandoc.table(tbl,keep.line.breaks=TRUE,caption=NULL,style="grid",split.cells=Inf,split.tables=Inf,justify='left',missing=''))
  table_str <- table_str[table_str != ""]
  cat(paste0(table_str,collapse="\n"),"\n\n",caption,"\n\n",sep="")
}

@@

The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019]. Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongTerm2003]. In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zone or mesophotic reefs intensifies. To date, the majority of such efforts have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018]. However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010]. Here, we investigate the factors influencing the community structure of brachyuran crabs across a suite of anthropogenic and environmental gradients using the Hawaiian Archipelago as a natural laboratory.

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean and is the most isolated island chain on the planet. Typically regarded as two regions, the main Hawaiian Islands (MHI) include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) comprise the uninhabited islands, atolls, and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau that make up the Papahānaumokuākea Marine National Monument. The islands are subject to considerable anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which are widely believed to influence coral reef communities. These gradients include differences in temperature, human population, and anthropogenic impact, among others. Studies examining community structure across broad environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalScaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007a]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. We then sampled brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [289 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011] from these structures to examine drivers of community composition across these gradients. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.

Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974]. While there are no shortages of exceptions, a long history of studies point towards a generalization that the lower bathymetric limits are generally regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001]. However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockRecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017]. While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalScaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].

One aspect of coral reef community structure that is thought to be important but remains under-investigatedis depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazMesophoticCoralBleaching2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021]. Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow crab communities vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow crab communities differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and anthropogenic factors drive crab community structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details]. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified (to species where possible, otherwise morphospecies), and preserved in 95% ethanol.<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->

>>>>

title: 'Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago'

@@

@@

  - himb: Hawai&#x02BB;i Institute of Marine Biology, University of Hawai&#x02BB;i, Mānoa

@@

  Coral reef communities are influenced by various environmental and anthropogenic factors, yet those factors have not been consistently resolved. Furthermore, due to logistical difficulties, investigations of the role of depth (particularly mesophotic coral ecosystems) in influencing coral reef community structure have been taxonomically limited. Here, we examined the effects of a suite of environmental, ecological, and anthropogenic factors on the assemblage structure of shallow brachyuran crabs across the ~2,500 km span of the Hawaiian Archipelago. In addition, we compared shallow crab assemblages with those sampled on a depth transect from 12--90 m on O&#x02BB;ahu. We deployed Autonomous Reef Monitoring Structures (ARMS) on shallow reefs (12--16 m) across the latitudinal gradient of the Hawaiian Archipelago to compare with a depth gradient (12, 30, 60, and 90 m) on the south shore of O&#x02BB;ahu. We found a total of 2,284 individual crabs representing 103 (morpho)species in 60 genera and 22 families. Shallow crab assemblages were significantly different between the Main and Northwest Hawaiian Islands and variation in community structure was significantly associated with six biotic, environmental, and anthropogenic factors (cumulative human impacts, chlorophyll-A, sea-surface temperature, site depth, island slope, and potential larval connectivity). In depth comparisons, we found more variation among mesophotic samples from O&#x02BB;ahu (30--90 m) than among all shallow samples across the complete span of the archipelago. We show that 90 m of depth is a more potent determinant of brachyuran assemblage composition than the latitudinal, environmental, or anthropogenic gradients across the Hawaiian Islands.<!--Deeper reefs host significantly different brachyuran communities than shallow ones in Hawai&#x02BB;i, and only 3 of 90 morphospecies (~3%) occurring across the entire depth range (mesophotic to shallow) were sampled. Further research on the cryptic reef-dwelling fauna is needed for identifying areas of unique biodiversity, as well as boundaries between depth-stratified reef ecosystems. Shallow coral reefs are extensively studied, and although scleractinian corals have been recorded to ~170 m, little is known about other mesophotic coral reef ecosystem (MCE) inhabitants. Brachyuran crabs fill many ecological and trophic niches on reefs, making them ideal candidates for evaluating species composition among depths. Here we ask if MCEs host the same communities as the shallower reefs spread across the entire 2400 km Hawaiian Archipelago.-->
bibliography: references/citations.json

@@

print_table <- function(tbl,id="",caption="",hdr_case="sentence",hide=character(0),accuracy = 0.01,format_numbers=TRUE, cell_width = Inf) {
  if (!grepl("^: ",caption)) {
    caption <- paste0(": ",caption)
  }

@@

  table_str <- pander::pandoc.table.return(
    tbl,
    keep.line.breaks=TRUE,
    caption=NULL,
    style="grid",
    split.cells=cell_width,
    split.tables=Inf,
    justify='left',
    missing=''
  )
  table_str <- table_str[table_str != ""]
  cat(paste0(table_str,collapse="\n"),"\n\n",caption,"\n\n",sep="")
}

@@

The structure and composition of coral reef communities is influenced by a suite of environmental and anthropogenic factors, yet the nature and makeup of those factors remain poorly resolved [@hughesCommunityStructureDiversity1989; @smithEffectsTopBottom2010; @brandlCoralReefEcosystem2019]. Coral reefs are among the most diverse habitats in the world, and they face increasing threats of degradation, including climate-driven bleaching and phase shifts to alternate ecological regimes [@grahamManagingResilienceReverse2013; @norstromAlternativeStatesCoral2009; @pandolfiGlobalTrajectoriesLongterm2003]. In the face of these threats, understanding drivers of metazoan community structure in coral reef ecosystems is critically important [@hughesRisingChallengeSustaining2010; @pageSeekingResistanceCoral2019], particularly as the search for criteria to prioritize climate refugia such as upwelling zone or mesophotic reefs intensifies. To date, the majority of such efforts have focused on conspicuous taxa such as corals and fishes [e.g., @benfieldComparisonCoralReef2008; @caballero-aragonWaveExposureTemperature2023; @kaneTrophicDesignationLive2017; @langeWaveExposureShapes2021; @vanlierHabitatConnectivityComplexity2018]. However, these taxa make up a minority of coral reef biodiversity and thus may not be indicative of community patterns overall [@fautinOverviewMarineBiodiversity2010]. Here, we investigate the factors influencing the community structure of brachyuran crabs across a suite of anthropogenic and environmental gradients using the Hawaiian Archipelago as a natural laboratory.

The Hawaiian Archipelago comprises high volcanic islands and atolls stretching across ~2,500 km in the North Pacific Ocean and is the most isolated island chain on the planet. Typically regarded as two regions, the main Hawaiian Islands (MHI) include the eight inhabited high islands, and the Northwestern Hawaiian Islands (NWHI) comprise the uninhabited islands, atolls, and banks northwest of Kaua&#x02BB;i and Ni&#x02BB;ihau. The NWHI are wholly enclosed by the Papahānaumokuākea Marine National Monument, one of the largest marine protected areas in the world. The islands are subject to considerable anthropogenic and physical gradients across their span [@selkoeMapHumanImpacts2009], which are widely believed to influence coral reef communities. These gradients include differences in temperature, coral cover, human population, and anthropogenic impact, among others. Studies examining community structure across broad environmental gradients have been criticized for confounding multiple potentially covarying factors (such as habitat area and/or complexity) across the measured gradients [e.g., @bellwoodRegionalscaleAssemblyRules2001; @grahamEffectsMarineReserve2003; @connollyCommunityStructureCorals2005; @ewersConfoundingFactorsDetection2006; @fletcherRoleHabitatArea2007]. Thus, we used autonomous reef monitoring structures (ARMS)---standardized collection devices designed to mimic reef structural complexity---to sample cryptic communities while controlling for habitat [@zimmermanArtificialReefMatrix2004; @brainardAutonomousReefMonitoring2009]. We then sampled brachyuran crabs, a group comprising a notable proportion of the coral reef cryptofauna [284 species in Hawai&#x02BB;i, @castroCatalogAnomuranBrachyuran2011] from these structures to examine drivers of community composition across these gradients. We compared crab communities among islands and regions (MHI vs. NWHI) and examined the effects of a suite of possible environmental and anthropogenic drivers of community structure to test hypotheses about their relative importance.

Much of our understanding of biotic zonation and community structure in the ocean originates with classic studies from the intertidal [e.g., @connellInfluenceInterspecificCompetition1961; @connellEffectsCompetitionPredation1961; @paineIntertidalCommunityStructure1974]. While there are no shortages of exceptions, a long history of studies point towards a generalization that the lower bathymetric limits are generally regulated by ecological interactions whereas upper limits of intertidal communities are more often determined by abiotic factors [@stephenson1972life; @mengeRockyIntertidalCommunities2001; @bertnessTestingRelativeContribution1999; @dennyPhysicalProcessesThat2001]. However, most of these models describe zonation at specific localities rather than community variation over larger spatial scales [but see @birdTideWaveMarks2013; @bishop-taylorTidesModellingElevation2019] and may not apply to coral reefs, which are subject to more complex community dynamics [@abramsRecruitmentLotteriesCoexistence1984; @bussCompetitiveNetworksNontransitive1979; @hixonPredationPreyRefuges1993; @hubbellUnifiedNeutralTheory2011; @moraEcologyFishesCoral2015; @saleStockrecruitRelationshipsRegional1982; @sheppardBiologyCoralReefs2017]. While there is little doubt that ecological interactions are important on coral reefs, the suite of factors and which play the most important role in structuring coral reef communities remains a subject of considerable debate [@bellwoodRegionalscaleAssemblyRules2001; @connollyUnifiedModelExplains2017; @hubbellUnifiedNeutralTheory2011; @jacksonAdaptationDiversityReef1991; @mumbyCoralReefHabitats2008; @saleMaintenanceHighDiversity1977].

One aspect of coral reef community structure that is thought to be important but remains under-investigated is depth. Most studies to date have focused on the upper ~30 m of the water column, yet coral reefs and associated communities are known to extend beyond 150 m [@pyleExploringDeepCoral1996; @pyleMesophoticCoralEcosystems2019], with photosymbiotic corals reported to 170 m [@rouzeSymbioticAssociationsDeepest2021]. As much as two thirds of coral reef ecosystems lie in this under-explored zone [@hindersteinThemeSectionMesophotic2010]. These habitats, known as mesophotic coral ecosystems (MCEs), are typically defined as those extending from ~30--150 m depth [@hindersteinThemeSectionMesophotic2010; @kahngCommunityEcologyMesophotic2010]. They have been hypothesized as possible refugia from climate change impacts on shallow reefs [@glynnCoralReefBleaching1996; @bongaertsAssessingDeepReef2010; @smithDepthRefugiumCatastrophic2014; @holsteinFertileFathomsDeep2015; @muirSpeciesIdentityDepth2017], although a growing number of studies have found evidence that mesophotic reefs are unique and uniquely threatened, challenging this notion [@diazLightTemperatureDrive2023; @diazMesophoticCoralBleaching2023; @hobanPlumbingDepthsEnvironmental2023; @hurleyAssessmentShallowMesophotic2016; @kaneHighLevelsMesophotic2014; @pinheiroDeepReefFishes2019; @rochaMesophoticCoralEcosystems2018; @montgomeryCommunitySimilaritySpecies2021]. Given the strong reported differences between deep and shallow reefs across taxa and the paucity of research on drivers of cryptic coral reef community variation, we sought to answer the following four research questions: (i) How do shallow reef brachyuran crab assemblages vary along the natural and anthropogenic gradients of the Hawaiian Archipelago? (ii) Do shallow reef brachyuran crab assemblages differ between the Main and Northwestern Hawaiian Islands? (iii) What environmental and anthropogenic factors drive crab assemblage structure in shallow coral reefs across the Hawaiian Archipelago? (iv) What are the relative contributions of depth (shallow vs. mesophotic) and island region to crab community variation across the Hawaiian Archipelago?

@@

ARMS units were retrieved by encapsulating them in-situ in a 100 μm mesh-lined crate to prevent escape of motile organisms. Once returned to the surface, they were placed in containers of aerated, filtered seawater, crates were removed, and units were systematically disassembled plate-by-plate [see @lerayDNABarcodingMetabarcoding2015 for processing details]. All adult brachyuran crabs ($\geq$ 5 mm) were photographed, identified, and preserved in 95% ethanol. Identifications were made to species level where possible, otherwise crabs were classified into morphospecies represented by suites of shared morphological characters.<!-- Specimens were collected under permits #xxx (NWHI) and #xxx (MHI) and accessioned into the invertebrate zoology collections at the Florida Museum of Natural History (catalog numbers xxx--xxx).-->
